---
title: "Class19_HW"
author: "Gabrielle Meza (A13747395)"
date: "12/1/2021"
output:
  pdf_document: default
  html_document: default
---

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
data <- read.table("Expression_Results.txt")
head(data)
```

```{r}
sum(data$geno == "A/G")
sum(data$geno == "G/G")
sum(data$geno == "A/A")
```


```{r}
library(dplyr)
ag <- data %>% filter(geno == "A/G")
gg <- data %>% filter(geno == "G/G")
aa <- data %>% filter(geno == "A/A")

```

```{r}
median(ag$exp)
median(gg$exp)
median(aa$exp)
```

A/G genotype: 233
A/G median expression: 25.06

G/G genotype: 121
G/G median expression: 20.07

A/A genotype: 108
A/A median expression: 31.24

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?


```{r}
boxplot(aa$exp, gg$exp, ag$exp)
boxplot(aa$exp, gg$exp, ag$exp, col= c("powderblue","purple","pink"), names = c("A/A", "G/G", "A/G"), xlab = "Genotype", ylab= "Expression")
```

People with the A/A genotype have higher ORMDL3 expression than the G/G genotype. This, along with the fact that A/G also is also has higher expression, is good evidence that the SNP of an A instead of G does have an impact on ORMDL3 expression, specifically an increase in expression. So these A/A genotypes, in my opinion, are more likely to have asthma. 
