---
title: "Class15 RNASeq Analysis"
author: "Gabrielle Meza (A13747395)"
date: "11/17/2021"
output: github_document
---

#Background

Today we're examining a published RNA-seq experiment where airway smooth muscle cells were yreated with dexamethasone, a synthetic glucocorticoid steriod with anti-inflammatory effects (Himes et al.)

We need:
1) Count of Data
2) col of data 

The count of data, is the actual data, with each gene as a column header. the col is the metadata, and each gene is a row header

```{r}
counts <- read.csv("airway_scaledcounts.csv", stringsAsFactors = FALSE, row.names=1)
metadata <-  read.csv("airway_metadata.csv", stringsAsFactors = FALSE)
```

```{r}
head(counts)
```
```{r}
head(metadata)
```

> Q1. How many genes are in this dataset? 

38694 total genes

```{r}
nrow(counts)
```


> Q2. How many ‘control’ cell lines do we have?

4

```{r}
sum(metadata$dex == "control")
```


Side note: 

lets check the correspondence of the metadata and count data setup 

```{r}
metadata $id
colnames(counts)
```

We can use the `==` thing to see if they colums and rows are the same 

```{r}
metadata$id == colnames(counts)
```

We are going to wrap this in an `all` thing so it will tell us if they all are true or not. if there is a flase andwhere, then it would not give out TRUE. We would use this for a larger dataset potentially.

```{r}
all(metadata$id == colnames(counts))
```


## Compare control to treated 

first we need to access all the control columns in our counts data. 

```{r}
metadata$dex == "control"
control.inds <- metadata$dex == "control"

#this is pulling out the rows that are control in the metadata sheet 
metadata[ control.inds, ]

#now use $id to get the ids that are controls and make it a new value 

control.ids <- metadata[ control.inds, ]$id

```

use these ids to access just the control columns of our `counts` data. Use head becuase it would be a large dataset 

```{r}
head(counts[ , control.ids])
```

```{r}
control.mean <- rowMeans(counts[ , control.ids])
head(control.mean)
```

> Q4. Follow the same procedure for the treated samples

```{r}
treated.inds <- metadata$dex == "treated"
treated.ids <- metadata[ treated.inds, ]$id
treated.mean <- rowMeans(counts[ , treated.ids])
head(treated.mean)
```

we will combine our meancount data for bookkeeping purposes

```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

there are `r nrow(counts)` rows/genes in this dataset. You can do this int he printout and have it run the code in the printout with *``* but you would have r rcomand in the *``*

## Compare the control and treated 

> Q5. Create a scatter plot showing the mean of the treated samples against the mean of the control samples. 

Quick plot of our progress so far

```{r}
plot(meancounts)

```

there should be many more points showen (38,000!) this could better be represented by a log scale 

```{r}
plot(meancounts, log="xy")
```

we often use log transformations as they make life much nicer in this world... 

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)

```

We need to drop the zero count genes/ rows! 

```{r}
head(meancounts[ ,1:2] ==0)
```

The `which()` function tells us the indices of TRUE entries in a logical vector. 

```{r}
which (c(T,F,T))
```

However, it is not that useful in default mode on our type of multi- column input 

```{r}
inds <- which(meancounts[,1:2] == 0, arr.ind=TRUE)
head(inds)
```

I only care about the rows here )if there is a zero in any column I will exclude this row eventually). You then want to use `unique` to get the rows you want to look at 

```{r}
to.rm <- unique(inds[,1])
mycounts <- (meancounts[-to.rm,])
```

```{r}
head(meancounts[-to.rm,])
```

we now have `r nrow(mycounts)` genes remaining. 

```{r}
nrow(mycounts)
```

How many of these genens are upregulated at the log 2 fold change threshold of +2 or greater?

```{r}
sum(mycounts$log2fc > +2)
```

what percentage is this?

```{r}
round((sum(mycounts$log2fc > +2) / nrow(mycounts))*100, 2)
```

How about the downreguolated genes? 

```{r}
sum(mycounts < -2)
```

## DESeq2 analysis 

```{r}
library(DESeq2)
```
We first need to setip the DESeq 

design is where in the col data do we care about? 

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
```

Run the DESeq anaylsis pipeline. it is doing a lot of other analysis, like p-value and stuff. it is being saved as a dds with all of the results. you can use the package default `results()` to read the data and out it into a new vairable to then to pull it up and read it all. 

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
head(res)
```

# A volcano plot 

This is a very common data viz of this type of data that does not really look like a volcano. you can use base r package to plot this

```{r}
plot(res$log2FoldChange, log(res$padj))
```

when you take a log, the values you are interested in at the very low numbers since it is a p-vaule. To make this easier visually, you can put a negative sign infront of log 

with log, as you move up in expression, go more positive. as you go down in expression, go more negative 

use abline to add a line at certain points. here we are making the line at the set pValue (horizontal) to be sugnificant. And then the cut off if there is a 2 fold change in either neg or positive direction. 

```{r}
plot(res$log2FoldChange, -log(res$padj))
abline(v=c(-2,2), col="gray")
abline(h=-log(0.05), col="gray")
```

## Adding annotation data 

Now lets color and annotate this data further! we want to add meaningful gene names to our dataset so we can make some sense of what is going on here! 

For this we will be installing and using 2 more packages in biocManager. I installed these in the console. **AnnotationDbi** the other contains data we are going to map between and is **org.Hs.eg.db**

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```
Here we map to "SYMBOL" the comon gene name that the world understands and wants 

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     keytype="ENSEMBL",
                     column="SYMBOL",   
                     multiVals="first")
head(res)
```

# Let's now save this data 

Let's finally save our results as a .csv file so we can use for a lter time, and use in different area if needed. 

```{r}
write.csv(res, file = "allmyresults.csv")
```


## DAY 2- Pathway analysis

Let's try to bring some biology insight back into this work. for this we will us KEGG. So let's load up KEGG for pathway analysis.

For this, we will have to install some stuff BiocManager::install( c("pathview", "gage", "gageData") )

```{r}
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)
```

```{r}
head(kegg.sets.hs, 2)
```

Before we can use KEGG we need to get our gen identifiers in the correct format for KEGG, which is ENTREZ format in this case. So let slook into the dataset again to get familiar with th elanguage 

```{r}
head(rownames(res))
```

```{r}
columns(org.Hs.eg.db)

```
The main **gage()** function requires a named vector of fold changes, where the names of the values are the Entrez gene IDs.

Note that we used the mapIDs() function above to obtain Entrez gene IDs (stored in res$entrez) and we have the fold change results from DESeq2 analysis (stored in res$log2FoldChange).

 
```{r}
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     keytype="ENSEMBL",
                     column="ENTREZID",   
                     multiVals="first")
res$genename <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     keytype="ENSEMBL",
                     column="GENENAME",   
                     multiVals="first")
```

Assign names to this vector that are the gene IDs that KEGG wants 

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

There we go! now let’s run the **gage()** pathway analysis. Get the results by doing keggres = and the gage function, and it saves the data as keggres

We can then look at the attributes() of this or indeed any R object

```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs)
attributes(keggres)
```

```{r}
head(keggres$greater)
```

```{r}
head(res)
```

The pathview() function will add our genes to a KEGG pathway as colored entries: 

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa05310")
```


