---
title: "Class17"
author: "Gabrielle Meza (A13747395)"
date: "11/24/2021"
output: github_document
  
---

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

```{r}
colnames(vax)
```

> Q1. What column details the total number of people fully vaccinated? 

"persons_fully_vaccinated"

>Q2. What column details the Zip code tabulation area? 

"zip_code_tabulation_area"

>Q3. What is the earliest date in this dataset? 

```{r}
min(vax$as_of_date)
```

>Q4. What is the latest date in this dataset? 

```{r}
max(vax$as_of_date)
```

let’s call the **skim()** function from the skimr package to get a quick overview of this dataset:

```{r}
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset? 

9 

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

```{r}
sum( is.na(vax$persons_fully_vaccinated) )
```

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
nrow(vax)
(8355/ 82908) * 100
```

One of the “character” columns of the data is as_of_date, which contains dates in the Year-Month-Day format.

Dates and times can be annoying to work with at the best of times. However, in R we have the excellent lubridate package, which can make life allot easier. Here is a quick example to get you started:

```{r}
library(lubridate)
today()
```

> Q9. How many days have passed since the last update of the dataset? 

```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
today() - vax$as_of_date[1]
```

In R we can use the **zipcodeR** package to make working with these codes easier

```{r}
library(zipcodeR)
geocode_zip('92037')
```

Calculate the distance between the centroids of any two ZIP codes in miles, e.g.

```{r}
zip_distance('92037','92109')
```

More usefully, we can pull census data about ZIP code areas (including median household income etc.

```{r}
reverse_zipcode(c('92037', "92109") )
```
> How many unique zip code

```{r}
length(unique(vax$zip_code_tabulation_area))
```

Subsetting can get tedious and complicated quickly. we will use the **filter()** function to do our subsetting from now on. this uses **dlypr()**

We want to focus in on the San Diego County 

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")
```

```{r}
nrow(sd)
```

> Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
length(unique(sd$zip_code_tabulation_area))
```
> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 

```{r}
which.max(vax$age12_plus_population)
```

```{r}
vax$zip_code_tabulation_area[which.max(vax$age12_plus_population)]
```

```{r}
sd$zip_code_tabulation_area[which.max(sd$age12_plus_population)]
```

More complicated subsetting

```{r}
sd.20 <- filter(vax, county == "San Diego", 
       age5_plus_population > 20000)

nrow(sd.20)
```

> Q13. what is the average vaccination rate of San Diego as of yesterday?

```{r}
sd.now <- filter(vax, county=="San Diego",
                 as_of_date=="2021-11-23")
head(sd.now)

sd.now$percent_of_population_fully_vaccinated
```

```{r}
summary(sd.now$percent_of_population_fully_vaccinated)
```

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”? As a histogram

```{r}
hist(sd.now$percent_of_population_fully_vaccinated)
```

This plot above is going to be susceptible to being skewed by ZIP code areas with small populations. these will have big effects for just a small number of unvax-ed folks... 

Now focus on UCSD/La Jolla area

> Q. What is the population of the 92037 ZIP code area?

```{r}
ucsd <- filter(sd.now, zip_code_tabulation_area=="92037")
ucsd$age5_plus_population
```

> Q. What is the average vaccination value for this UCSD/La Jolla ZIP code area?

```{r}
ucsd$percent_of_population_fully_vaccinated
```
 
 > Lets do my zipcode! 92124 and then where i am going: 92065
 
```{r}
tierrasanta<- filter(sd.now, zip_code_tabulation_area=="92124")
tierrasanta$age5_plus_population
tierrasanta$percent_of_population_fully_vaccinated

```
```{r}
ramona <- filter(sd.now, zip_code_tabulation_area=="92065")
ramona$percent_of_population_fully_vaccinated
```
 
Lets make a time series of vacination rate for a given ZIP code area. I will do 92124

```{r}
ttown <- filter(vax, zip_code_tabulation_area == "92124")
library(ggplot2)
```

```{r}
ggplot(ttown) + 
  aes(x=as_of_date, y=percent_of_population_fully_vaccinated) + 
  geom_point() +
  labs(x="Date", y="Percent Vaccinated")
```

Let's make this plot for all of SD county zip codes that have a population as large as 92037 (UCSD)

```{r}
sd.36 <- filter(vax, county=="San Diego",
                age5_plus_population > 36144)
head(sd.36)
```

> Lets do a plot for all of california, with similar populations. Populations bigger than UCSD: 36144

```{r}
ca <- filter(vax, age5_plus_population > 36144)
```

How many zip codes withth is pop?

```{r}
length(unique(ca$zip_code_tabulation_area))
```

Now lets make the plot

```{r}
ggplot(ca) + 
  aes(x=as_of_date, y=percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpjha=0.2) +
  labs(x="Date", y="Percent Vaccinated")
```

What is the mean across the state for these 36k+ population areas? 

```{r}
ca.now <- filter(ca, as_of_date=="2021-11-22")
summary(ca.now$percent_of_population_fully_vaccinated)
```

Add a line for the mean % of people vaccinated in California. 

```{r}
 ggplot(ca) + 
  aes(x=as_of_date, y=percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpjha=0.2) +
  labs(x="Date", y="Percent Vaccinated") +
  geom_hline(yintercept = 0.67, color = "Red")
```





